Temporality of the Victorian Hospital Patient Transfer Network

### 2025-09-01 reports/paper.qmd
? model rewrite

We use admissions data sourced from the Victorian Admitted Episdoes Dataset (VAED) [@VAED_official]. This contains line-listed admissions for patients in Victoria, Australia, covering 338 healthcare facilities that vary in size, case mix and services.
the dataset contains 26,796,407 individual admissions across 5,471893 unique patients, from 1 January 2011 through to 6 December 2019 (3631 days).
Overlapping admissions for a patient are resolved by deferring to the patient's most recent admission event, and the method is described in detail in the Appendix, and is used in [@donker_patient_2010,@nekkab_assessing_2020].

Some exploratory analysis of this data was done in [@lydeamore_carbapenemase-producing_2024] using a static network framework, and it characterises the structure of the healthcare system.

<!--# Model description starts here -->

Admissions data can be used to inform the construction of a wide variety of (weighted, directed) networks, which can, in turn, be used to construct a stochastic process model. In this study, we use discrete time Markov chain models, where a state vector $x$ that represents the number of marked individuals at facilities $u$. Weights $w_{ij}$ of the edges $e_{ij}$ between nodes $u_i, u_j$ of the network inform the hazard function of the movement process of marked individuals between those nodes.
By default, the movement process is a Markovian process, i.e. is a Poisson random variable with hazard $\lambda_{ij}(x_i, t) = c_iw_{ij}x_i\Delta t$, where $c_i$ is a normalising constant, and $\Delta t$ is the time step.

Applying this directly to a static network of transfers between healthcare facilities, with appropriate infection dynamics, is what is done in @donker_hospital_2012, for example, and represents a model where ....
However, since individuals tend to leave the hospital system for non-negligible amoutns of time, it may be appropriate to include time spent in the community (e.g. at home) as additional nodes of the network. 
In @belik_leveraging_2017, this approach is taken with a temporal network approach in snapshot representation, where the system over a time period is represented as a series of static networks that "snapshot" the system at discrete points in time. 
Through examination of the behaviour of transfers, especially those that are indirect and cross-facilitly, we want to expand on this temporal representation, and include edges that explicityl ...

We can capture the This motivates the general representation in @fig-general-diagram, where edges from a location $u$ can go to null (discharged and never seen again), directly to some location $v$ (that could be $u$), or indirectly to $v$ via a community node $h_{uv}$.

![Sketch of the generalised model with processes of (1) permanent discharge ($u \to \emptyset$), (2) direct transfer ($u \to v$), (3) return home ($u \to h_{uv}$), (4) readmission $h_{uv} \to v$).](diagrams/movement_general_case.pdf){#fig-general-diagram}

The governing equations for such models can be captured by sets of transitions modelled as a Markov chain. The hazards of the transition events is given in @eq-general-model-transitions.

$$
\begin{aligned}
\lambda(u(s) \to \emptyset) &= \zeta_u(s)\\
\lambda(u(s) \to v(t)) &= d_{uv}(s, t)\\
\lambda(u(s) \to z_{uv}(s, t)) &= \eta_{uv}(s, t)\\
\lambda(z_{uv}(s, t) \to v(t)) &= \rho_{uv}(s, t)
\end{aligned}
$$ {#eq-general-model-transitions}

In order in increase the flexibility of the models, we will use a _discrete time_ Markov chain, where we have a Poisson-distributed number of transition events occur within some given timestep. For some models, we will also allow for the number of transitions for a particular class of event to be non-Poisson.

<!--# Model abstraction ends here, model details start here -->

@@

$$ d_{uv}(s,t) = d_{uv} = \frac{\sum_{s,t} w_{uv}(s,t)}{T_\Sigma N_u} $$
$$ \eta_{uv}(s,t) = \eta_{uv} = 0 $$
$$ \rho_{uv}(s,t) = \rho_{uv} = 0 $$

@@

$$d_{uv}(s,t) = d_{uv} =  \frac{\sum_{s,t: (t-s) < \omega} w_{uv}(s,t)}{T_\Sigma N_u}$$
$$\eta_{uv}(s,t) = \eta_{uv} = \frac{\sum_{s,t: (t-s) \geq \omega} w_{uv}(s,t)}{T_\Sigma N_u}$$
$$\rho_{uv}(s,t) = \rho_{uv} = \left[ \frac{\sum_{s,t: (t-s) \geq \omega} (t-s) w_{uv}(s,t)}{\sum_{s,t: (t-s) \geq \omega} w_{uv}(s,t)}  \right]^{-1}$$

@@

$$d_{uv}(s,t) = d_{uv}(s) = \frac{\sum_{t: (t-s) < \omega} w_{uv}(s,t)}{\omega N_u}$$
$$\eta_{uv}(s,t) = \eta_{uv}(s) = \frac{\sum_{t: (t-s) \geq \omega} {w_{uv}(s, t)}}{\omega N_u}$$
$$\rho_{uv}(s,t) = \rho_{uv}(t) = \frac{\sum_{s: (t-s) \geq \omega} {w_{uv}(s, t)}}{\omega H_{uv}}$$

@@

$$d_{uv}(s, t) = \frac{\sum_{t: (t-s) < \omega} w_{uv}(s, t)}{\omega N_u}$$
$$\eta_{uv}(s, t) = \frac{\sum_{t: (t-s) \geq \omega} w_{uv}(s, t)}{\omega N_u}$$
$$\rho_{uv}(s,t) = \frac{1}{\lceil{}t\rceil{}_\omega - t}$$

@@

In the current study, we have used a discrete time stepping scheme in order to allow for some non-Markovian behaviour; it is likely that an agent-based model would be a more "accurate" representation. 
We have also neglected "higher-order" pathing. [@myall_network_2021]

>>>>

We use admissions data sourced from the Victorian Admitted Episdoes Dataset (VAED) [@VAED_official]. This contains line-listed admissions for patients in Victoria, Australia, covering 338 healthcare facilities that vary in size, case mix and services.
the dataset contains 26,796,407 individual admissions across 5,471893 unique patients, from 1 January 2011 through to 6 December 2019 (3631 days).
Some exploratory analysis of this data was done in [@lydeamore_carbapenemase-producing_2024] using a static network framework, and it characterises the structure of the healthcare system.
Overlapping admissions for a patient are resolved by deferring to the patient's most recent admission event, and the method is described in detail in the Appendix, and is used in [@donker_patient_2010,@nekkab_assessing_2020].

As stated above, existing literature does examine patient transfer networks using network analysis methods. These encompass a large range of dynamics and temporal fidelity of movement.
Here, we will consider a number of different models of movement at various levels of temporal fidelity. In general, we will decompose the model of patient movement as in @fig-general-diagram, where a patient is discharged at location $u$ at time $s$, to either never readmit (move to $\emptyset$), directly transfer to location $v$ at time $s$, or return home $h_{uv}$ and admit at location $v$ at some later time $t$. This dynamic is replicated at each pair of locations $(u, v)$, and potentially over all pairs of times $(s, t)$.
The hazards for these processes are used to define discrete-time stochastic processes that can be analysed, and will be simulated, in order to understand the dynamics of the movement models.
These can be captured in @eq-general-model-transitions:

$$
\begin{aligned}
\lambda(u(s) \to \emptyset) &= \zeta_u(s)\\
\lambda(u(s) \to v(s)) &= d_{uv}(s)\\
\lambda(u(s) \to h_{uv}(s, t)) &= \eta_{uv}(s, t)\\
\lambda(h_{uv}(s, t) \to v(t)) &= \rho_{uv}(s, t)
\end{aligned}
$$ {#eq-general-model-transitions}

@@

$$ d_{uv}(s) = d_{uv} = \frac{\sum_{s,t} w_{uv}(s,t)}{T_\Sigma N_u} $$
$$ \eta_{uv}(s,t) = \eta_{uv} = 0 $$
$$ \rho_{uv}(s,t) = \rho_{uv} = 0 $$

@@

$$d_{uv}(s) = d_{uv} =  \frac{\sum_{s,t: (t-s) < \omega} w_{uv}(s,t)}{T_\Sigma N_u}$$
$$\eta_{uv}(s,t) = \eta_{uv} = \frac{\sum_{s,t: (t-s) \geq \omega} w_{uv}(s,t)}{T_\Sigma N_u}$$
$$\rho_{uv}(s,t) = \rho_{uv} = \left[ \frac{\sum_{s,t: (t-s) \geq \omega} (t-s) w_{uv}(s,t)}{\sum_{s,t: (t-s) \geq \omega} w_{uv}(s,t)}  \right]^{-1}$$

@@

$$d_{uv}(s) = \frac{\sum_{t: (t-s) < \omega} w_{uv}(s,t)}{\omega N_u}$$
$$\eta_{uv}(s,t) = \eta_{uv}(s) = \frac{\sum_{t: (t-s) \geq \omega} {w_{uv}(s, t)}}{\omega N_u}$$
$$\rho_{uv}(s,t) = \rho_{uv}(t) = \frac{\sum_{s: (t-s) \geq \omega} {w_{uv}(s, t)}}{\omega H_{uv}}$$

@@

$$d_{uv}(s) = \frac{\sum_{t: (t-s) < \omega} w_{uv}(s, t)}{\omega N_u}$$
$$\eta_{uv}(s, t) = \frac{\sum_{t: (t-s) \geq \omega} w_{uv}(s, t)}{\omega N_u}$$
$$\rho_{uv}(s,t) = \frac{1}{\lceil{}t\rceil{}_\omega - t}$$

@@

There are some clear avenues of further investigation from the current study.
One of these is the addition of disease transmission dynamics. We chose to neglect these mechanisms due to the complexity of defining and characterising the manner of transmission within a facility -- and this can have confounding effects on the interpretation of observed results. It is believed that CPE, for example, is transmitted through the environment, with

In the current study, we have used a discrete time stepping scheme in order to allow for some non-Markovian behaviour; it is likely that an agent-based model would be a more "accurate" representation. This would allow for "higher-order" behaviours, such as those observed in [@myall_network_2021], where patient pathways through wards in that case, and hospitals in ours, are dependent on their admission history.
However, agent-based models are an order of magnitude more expensive to simulate, due to increased computational and memory requirements. The number of
